A Parallel General Purpose Multi-Objective Optimization Framework, with Application To Beam Dynamics

### 2018-07-09 paper/theor-experiment.tex
updating paper

%Currently we run the simulations on a low resolution space charge grid ($8
  %\times 8 \times 8$) as a proof of concept.
%In a next step a simple Python script would be applied to iteratively rerun
  %the optimizer with increasing space-charge resolution when the limits of the
  %design variables are narrow enough (iteratively diminishing search space). 
%\vspace{-7em}
\subsection{AWA Photoinjector Optimization} \label{awaproblem}
Next we apply the optimization framework to the high charge beam line at the AWA. 
As shown in Fig.  \ref{awa-pic}, this beam line consists of an rf photocathode gun, 
two solenoids, six linear accelerating cavities, 
followed by four quadrupoles and a stripline kicker. 
The charge of interest is 40 nC, and based on Two Beam 
Acceleration (TBA) \cite{gai_power_jing_2012} experiments performed in the past 
at AWA. Similar experiments will take place in the future and motivate this work. 
Prior experimental results were limited by beam size and emittance as the beam passed through small aperture 
wakefield structures located after the quadrupoles. 
In an attempt to maximize charge transmission in upcoming experiments
we optimize several parameters in the gun and quadrupoles leading into the 
TBA section of the beam line.
%\vspace{-1em}

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{Report/awa-drawing}
	\caption{High charge beam line layout at the AWA.}
	\label{awa-pic}
\end{figure}

%\vspace{-1em}
In addition to the physics challenges, we chose this model to demonstrate the frameworks ability to tackle large problems.
The number of design variables is eight, and the number of objectives is seven.
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.

\begin{align}
\text{min}  \quad & \text{rms}_{x}\left(s = s_1\right), \quad \text{rms}_{x}\left(s = s_2\right) \label{eq:awa:p1}\\
& \text{rms}_{s}\left(s = s_1\right), \quad \text{rms}_{s}\left(s = s_2\right), \label{eq:awa:p2}\\
& \varepsilon_{x}\left(s = s_1\right), \quad \varepsilon_{x}\left(s = s_2\right), \label{eq:awa:p3} \\
& dE \label{eq:awa:p4} \\
\text{subject to} \quad & q = 40 \left[\text{nC}\right] \label{eq:awa:firstconstr}\\
\quad & \text{Volt}_{\text{Gun}} = 64\left[\text{MV/m}\right] \label{eq:awa:lastconstr}\\
\quad & \text{Volt}_{\text{Linac}} = 24-25\left[\text{MV/m}\right] \\
\quad & \sigma_x = \sigma_y = 9 \left[\text{mm}\right] \label{eq:awa:firstdvar}\\
\quad & \phi_{\text{linac}} =0^\circ \label{eq:awa:lastdvar}
\end{align}

The first four objectives, Eqs. (\ref{eq:awa:p1}) and (\ref{eq:awa:p2}),
minimize the transverse ($rms_x$) and longitudinal ($rms_s$) 
beam sizes at two locations of interest in the beam line ($s_1$,$s_2$);
the longitudinal positions at the entrance of the kicker and 
$1$ meter after the exit of the kicker. 
Minimizing the beam size at these locations is essential to 
to preventing scraping and ensuring transmission through the wakefield structures 
following the kicker. The next two objectives (\ref{eq:awa:p3}) minimize the 
transverse emittance, which favors less divergence in the beam and reduces 
emittance growth during transport. The last objective (\ref{eq:awa:p4}), 
minimizes energy spread in the bunch at the entrance of the kicker at $s_1$.  
This helps reduce emittance and beam size growth in the kicker.  
Equations (\ref{eq:awa:firstconstr}) to
(\ref{eq:awa:lastdvar}) define the charge, gun voltage, linac voltages, 
laser radius and linac cavity phases (in that order). These are parameters in the simulation 
that must be defined, but do not vary during the optimization.

%\vspace{-1em}
As introduced in Section \ref{ferrario}, we define the AWA optimization problem 
objectives, and design variables in the OPAL input file as shown in
the following code: %equations (\ref{eq:awa:p1}) to (\ref{eq:awa:lastdvar}).

\vspace{0.2cm}
{\footnotesize \begin{verbatim}
	// Gun variables
	dv0: DVAR, VARIABLE="IBF",    LOWERBOUND=350.0, UPPERBOUND=500.0;
	dv1: DVAR, VARIABLE="IM",     LOWERBOUND=170.0, UPPERBOUND=260.0;
	dv2: DVAR, VARIABLE="GPHASE", LOWERBOUND=-30.0, UPPERBOUND=0.0;
	dv3: DVAR, VARIABLE="FWHM",   LOWERBOUND=1.5,   UPPERBOUND=10.0;

@@

	// Objectives
	rmss1:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_s',16.45))";
	rmsx1:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_x',16.45))";
	emitx1: OBJECTIVE,EXPR="fabs(statVariableAt('emit_x',16.45))";
	de1:    OBJECTIVE,EXPR="fabs(statVariableAt('dE',16.45))";	
	rmss2:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_s',18.45))";
	rmsx2:  OBJECTIVE,EXPR="fabs(statVariableAt('rms_x',18.45))";
	emitx2: OBJECTIVE,EXPR="fabs(statVariableAt('emit_x',18.45))";

@@

Design variables include the currents in two gun solenoids (IBF and IM), 
the gun phase (GPHASE), the laser longitudinal full width half max (FWHM), 
and four quadrupole strengths (KQ1-KQ2). The objectives include
beam size, emittance, and energy spread as
defined in Eqs. (\ref{eq:awa:p1}) to (\ref{eq:awa:p4}). 
The location at the entrance of the kicker is $s_1=16.45$~meters, 
and the location of interest after the kicker is $s_1=18.45$~meters.

All simulations for this experiment were carried out on Bebop a
high performance computing (HPC)
cluster provided by the Laboratory Computing Resource Center (LCRC)
at Argonne National Laboratory (ANL). We used Intel Knights Landing 
(KNL) processors at 1.3 GHz with 128 GB of memory 
and 64 cores per node. There are 352 compute nodes available on 
Bebop, with a total of 22,528 cores. All jobs were run and compared 
on 8 cores each, which allowed 8 jobs per node on the KNLs.
This in combination with the amount of nodes available 
allows for very large optimization jobs, like the AWA case.

A study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. 
The grid size was $16 \times 16 \times 32$, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step (dT) near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}. For all models, the longitudinal parameters (rms$_s$ and energy) 
are calculated correctly, but we see discrepancies in the transverse 
rms$_x$ and $\epsilon_x$ in the low fidelity results. 
Average run time each simulation with the adjusted time steps was 1.6~minutes.
In comparison the mid fidelity simulations ran for 18~minutes each.
%A summary of the parameters used is listed in Table \ref{fidelity}.

@@

\subsubsection{AWA Optimization Results}
While the optimization problem and goals were well defined (Section \ref{awaproblem}), 
it was not clear what the best input parameters for the NSGA-II should be.
These parameters include gene mutation probability, mutation probability, 
recombination probability, number of individuals (initial and and for later generations), 
and number of generations to complete. 
Given the beam line and design variables given in Section \ref{awaproblem}, 
four optimization experiments were done with varying NSGA-II parameters. 
Similar to the time step scan, 
the goal of this exercise was to determine which set of optimization
parameters strongly influence the results, 
and whether there was a time to solution difference.
From here on, we will reference each experiment as ex-1, ex-2, ex-3, and ex-4
as shown in Table \ref{extable}. 
\begin{table}%[h!]
	\begin{center}
		\caption{Input Parameters for initial twenty four hour AWA optimization experiments. 
			The gene mutation probability was equal to the mutation probability (not shown) in all four experiments. 
			The max number of individuals per generation was~80.}
		\label{extable}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C C D }
			\hline\noalign{\smallskip}
			& Gene Mutation Probability & Recombination Probibility & Number of completed generations \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			ex-1 &  0.1  & 0.9  &  96 \\
			ex-2 &  0.3  & 0.7  &  81 \\
			ex-3 &  0.8  & 0.2  &  53 \\
			ex-4 &  0.01 & 0.09 &  95 \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

The maximum number of individuals per generation was fixed at 80, 
Each experiment was allowed to run for twenty four hours, with 
a maximum generation limit of 100. 
We reduced the seven objectives to the four needed to construct 
the Pareto fronts that would be the final goal of this study.
The objectives include: $\varepsilon_{x}\left(s = s_1\right)\text{, } \varepsilon_{x}\left(s = s_2\right)\text{, } \text{rms}_{s}\left(s = s_1\right)\text{, and }  \text{rms}_{s}\left(s = s_2\right)$. 
The optimization parameters specific to OPAL input for ex-3 is given as an example.
Note the design variables defined in Section \ref{awaproblem}
precede the optimization parameters in the OPAL input file.

@@

The initial number of individuals was fixed at 656, 
and the minimum number individuals in later generations was fixed at 328. 
These numbers we in part based on the architecture of the KNL's. 
Since each simulation takes 8 cores, and there are 64 cores per KNL node, 
we wanted a large population size that would fit evenly on these resources.

>>>>

%Currently we run the simulations on a low resolution space charge grid ($8
  %\times 8 \times 8$) as a proof of concept.
%In a next step a simple Python script would be applied to iteratively rerun
  %the optimizer with increasing space-charge resolution when the limits of the
  %design variables are narrow enough (iteratively diminishing search space). 
%\vspace{-7em}
\subsection{AWA Photoinjector Optimization} \label{awaproblem}
Next we apply the optimization framework to the high charge beam line at the AWA. 
The purpose of the Argonne Wakefield Accelerator (AWA) optimization is to produce beams of electrons meeting 
design specifications including number of particles (charge), energy, and particle distribution (characterized by transverse beam sizes and energy spread).
As shown in Fig.  \ref{awa-pic}, this beam line consists of an rf photocathode gun, 
two solenoids, six linear accelerating cavities, 
followed by four quadrupoles and a stripline kicker. 
The charge of interest, 40 nC, is needed for Two Beam 
Acceleration (TBA) \cite{gai_power_jing_2012} experiments performed in the past 
at AWA. Similar experiments will take place in the future and motivate this work. 
Prior experimental results were limited by beam size as the beam passed through small aperture 
wakefield structures located after the quadrupoles. 
In an attempt to maximize charge transmission in upcoming experiments
we optimize several parameters in the gun and quadrupoles leading into the 
TBA section of the beam line.
%\vspace{-1em}

\begin{figure}
	\begin{tikzpicture}[every node/.style={anchor=south west,inner sep=0pt},x=1mm, y=1mm,]   
	\node (fig1) at (0,0)
	{\includegraphics[width=0.9\textwidth]{Report/awa-drawing}};
	\node[fill=white, inner sep=2pt] (txt2) at (35,15) {Interferometer};
	\node[fill=white, inner sep=2pt, rotate=26] (txt2) at (18,19.5) {Slit};	
	\node[fill=white, inner sep=2pt, rotate=20] (txt2) at (13,27) {Window};
	\end{tikzpicture}
	\caption{High charge beam line layout at the AWA.}
	\label{awa-pic}
\end{figure}
%\vspace{-1em}
In addition to the physics challenges, we chose this model to demonstrate the ability of the framework to tackle large problems.
Six design variables and objectives were used, 
along with three constraints.
The objectives include transverse and longitudinal beam size, 
transverse momentum and longitudinal energy spread. 
The design variables include the two gun solenoids and 
the first four quadrupoles. The objectives are calculated at 
$s_1=19.4$~[m], at the entrance of the fifth quadrupole in Fig.~\ref{awa-pic}.  
This problem encompasses high dimensionality 
and nonlinear effects such as space charge. There is no existing information
that accurately predicts where the Pareto front will fall. This work is also meaningful in that
it will guide future operations at the AWA.

\begin{align}
\text{min}  \quad & \text{rms}_{x}\left(s = s_1\right), \quad \text{rms}_{y}\left(s = s_1\right) \label{eq:awa:p1}\\
& \text{rms}_{px}\left(s = s_1\right), \quad \text{rms}_{py}\left(s = s_1\right), \label{eq:awa:p2}\\
& \text{rms}_{s}\left(s = s_1\right), \quad dE \label{eq:awa:p4} \\
\text{subject to} \quad & rms_x(s=s_0) < 0.1 \, [m]\label{eq:awa:c1}\\
\quad & rms_y(s=s_0) < 0.1\, [m]\label{eq:awa:c2}\\
\quad & |rms_y(s=s_0) - rms_x(s=s_0) | < 0.005 \, [m]\label{eq:awa:c3}\\
\quad & q = 40 \left[\text{nC}\right] \label{eq:awa:firstconstr}\\
\quad & \text{Volt}_{\text{Gun}} = 64\left[\text{MV/m}\right] \label{eq:awa:lastconstr}\\
\quad & \text{Volt}_{\text{Linac}} = 24-25\left[\text{MV/m}\right] \\
\quad & R_x = R_y = 9 \left[\text{mm}\right] \label{eq:awa:firstdvar}\\
\quad & \phi_{\text{gun}} =-20^\circ \label{eq:awa:gphidvar}\\
\quad & \phi_{\text{linac}} =-20^\circ \label{eq:awa:lastdvar}
\end{align}

The first four objectives, Eqs. (\ref{eq:awa:p1}) to (\ref{eq:awa:p2}),
minimize the transverse ($rms_{x,y}$) beam size and transverse momentum ($rms_{px,py}$)
at the location of interest in the beam line ($s_1$).
Where $s_1$ is the longitudinal entrance of the fifth quadrupole. 
Minimizing the beam size at this location is essential to 
to preventing loss of particles by scraping.
Preventing scraping ensures better transmission through the wakefield structures downstream. 
Less divergence in the beam (lower transverse momentum) 
reduces growth of transverse beam size during transport in bending elements, 
which also helps with transmission downstream. 
The next two objectives (\ref{eq:awa:p2}) minimize the 
longitudinal beam size ($rms_s$), and energy spread (dE) at $s_1$.
This helps reduce longitudinal beam size growth in bending elements.

Three constraints were also defined to help guide the algorithm;
Eqs.~\ref{eq:awa:c1} to \ref{eq:awa:c3}.
The bounding values may seem large, but this is done to avoid over constraing the 
problem in early generations. If the constraints are set tighter, 
runs do not reach the second generation, because enough satisfactory 
points are not produced to fill a complete generation. 
Leaving the value large allows for enough simulations to finish
in order to reach the next generation, and subsequent improved points in later generations.
The difference constraint, Eq.~\ref{eq:awa:c3}, is used to favor nearly round beams.
This prevents one dimension from becoming disproportionately large compared to the other.
In this case, there is some room in the beam pipe to allow the y dimension to grow, but round beams 
are preferred for easier data analysis.

Equations (\ref{eq:awa:firstconstr}) to
(\ref{eq:awa:lastdvar}) define the charge, gun voltage, linac voltages, 
laser radius, gun phase, and linac cavity phases (in that order). 
These are parameters in the simulation 
that must be defined, but do not vary during the optimization.
%\vspace{-1em}
As introduced in Section \ref{ferrario}, we define the AWA optimization 
design variables, objectives, and constraints in the OPAL input file as shown in
the following code: %equations (\ref{eq:awa:p1}) to (\ref{eq:awa:lastdvar}).

\vspace{0.2cm}
{\footnotesize \begin{verbatim}

	// Gun variables 
	dv0: DVAR, VARIABLE="IBF",    LOWERBOUND=300.0, UPPERBOUND=500.0;
	dv1: DVAR, VARIABLE="IM",     LOWERBOUND=180.0, UPPERBOUND=280.0;

@@

	//Objectives
	de3: OBJECTIVE,EXPR="fabs(statVariableAt('dE',19.4))";
	rmss3: OBJECTIVE,EXPR="fabs(statVariableAt('rms_s',19.4))";
	rmsx3: OBJECTIVE,EXPR="fabs(statVariableAt('rms_x',19.4))";
	rmsy3: OBJECTIVE,EXPR="fabs(statVariableAt('rms_y',19.4))";
	rmspx3: OBJECTIVE,EXPR="fabs(statVariableAt('rms_px',19.4))";
	rmspy3: OBJECTIVE,EXPR="fabs(statVariableAt('rms_py',19.4))";

	//Kicker apeture
	c1: CONSTRAINT, EXPR="fabs(statVariableAt('rms_x',16.5))<0.1";
	c2: CONSTRAINT, EXPR="fabs(statVariableAt('rms_y',16.5))<0.1";
	c3: CONSTRAINT, EXPR="fabs(statVariableAt('rms_y',16.5)-statVariableAt('rms_x',16.5))<0.005";

	OPTIMIZE, INPUT="tmpl/optLinac-40nC.tmpl",
	OUTPUT="optLinac-40nC",
	OUTDIR="results",
	OBJECTIVES = {rmss3, rmsx3, rmsy3, rmspx3, rmspy3, de3},
	DVARS = {dv0, dv1, dv4, dv5, dv6, dv7},
	CONSTRAINTS = {c1, c2, c3},
	INITIALPOPULATION=656,
	MAXGENERATIONS=200,
	NUM_MASTERS=1,
	NUM_COWORKERS=8,
	SIMTMPDIR="tmp",
	TEMPLATEDIR="tmpl",
	FIELDMAPDIR="fieldmaps",
	NUM_IND_GEN=328,
	GENE_MUTATION_PROBABILITY=0.01,
	MUTATION_PROBABILITY=0.01,
	RECOMBINATION_PROBABILITY=0.09;
	QUIT;

@@

Design variables include the currents in two gun solenoids (IBF and IM), 
and four quadrupole strengths (KQ1-KQ2). The objectives include
beam size (transverse and longitudinal), transverse momentum, and energy spread as
defined in Eqs. (\ref{eq:awa:p1}) to (\ref{eq:awa:p4}). 
The location at the entrance of the kicker is $s_0=16.45$~meters, 
and the location of interest for objectives is $s_1=19.4$~meters.

All simulations for this experiment were carried out on Bebop a
high performance computing (HPC)
cluster provided by the Laboratory Computing Resource Center (LCRC)
at Argonne National Laboratory (ANL). We used Intel Knights Landing 
(KNL) processors at 1.3 GHz with 128 GB of memory 
and 64 cores per node. There are 352 compute nodes available on 
Bebop, with a total of 22,528 cores. All jobs were run and compared 
on 8 cores each, which allowed 8 jobs per node on the KNLs.
This in combination with the number of nodes available 
allows for very large optimization jobs, like the AWA case.
Typical runs for this paper used 41 nodes, which corresponds to 2624 KNL cores 
and a generation size of 328 individuals.

A study on time step and 
number of particles was done to reduce the time to simulation while 
maintaining the physics of interest. 
The grid size $16 \times 16 \times 32$ was chosen, 
and parallelized in the x and y directions.
After comparing several options we choose ten thousand particles 
in combination with adjusting the time step (dT) near sensitive elements 
(i.e. quadurpoles, kicker). 
The resulting simulations are low fidelity, but closely approximate 
the mid fidelity simulations for metrics of interest, as shown in 
Fig. \ref{tstep}. For all models, the longitudinal parameters (rms$_s$ and energy) 
are calculated correctly, but we see discrepancies in the transverse 
rms$_x$ and $\epsilon_x$ in the low fidelity results. The
average run time of each simulation with the adjusted time steps was 1.6~minutes.
In comparison the mid fidelity simulations ran for 18~minutes.
%A summary of the parameters used is listed in Table \ref{fidelity}.

@@

\subsubsection{AWA Optimization Results}
While the optimization problem and goals were well defined (Section \ref{awaproblem}), 
it was not clear what the best input parameters for the NSGA-II should be.
These parameters include gene mutation probability, mutation probability, 
recombination probability, number of individuals (initial \lsnote{you have an extra `and' here} and for later generations), 
and number of generations to complete. 
Given the beam line and design variables given in Section \ref{awaproblem}, 
four optimization experiments were done with varying NSGA-II parameters. 
Similar to the time step scan, 
the goal of this exercise was to determine which set of optimization
parameters strongly influence the results, 
and whether there was a time to solution difference.
From here on, we will reference each experiment as ex-1, ex-2, ex-3, and ex-4
as shown in Table \ref{extable}. 
\begin{table}%[h!]
	\begin{center}
		\caption{Input Parameters for initial twenty four hour AWA optimization experiments. 
			The gene mutation probability was equal to the mutation probability (not shown) in all four experiments. 
			The max number of individuals per generation was~80.}
		\label{extable}
		\begin{tabular*}{\textwidth}{l @{\extracolsep{\fill}} C C D }
			\hline\noalign{\smallskip}
			& Gene Mutation Probability & Recombination Probibility & Number of completed generations \\
			\noalign{\smallskip}\hline\noalign{\smallskip}
			ex-1 &  0.1  & 0.9  &  96 \\
			ex-2 &  0.3  & 0.7  &  81 \\
			ex-3 &  0.8  & 0.2  &  53 \\
			ex-4 &  0.01 & 0.09 &  95 \\
			\noalign{\smallskip}\hline
		\end{tabular*}
	\end{center}
\end{table}

The maximum number of individuals per generation was fixed at 80, 
Each experiment was allowed to run for twenty four hours, with 
a maximum generation limit of 100. 
We reduced the seven objectives to the four needed to construct 
the Pareto fronts that would be the final goal of this study.
The objectives include: $\varepsilon_{x}\left(s = s_1\right)\text{, } \varepsilon_{x}\left(s = s_2\right)\text{, } \text{rms}_{s}\left(s = s_1\right)\text{, and }  \text{rms}_{s}\left(s = s_2\right)$. 
\lsnote{You could define $\varepsilon_{x1}$ to be $\varepsilon_{x} (s=s1)$ and similarly in the table and then use going forward to simplify notation}
The optimization parameters specific to OPAL input for ex-3 \lsnote{are} given as an example.
Note the design variables defined in Section \ref{awaproblem}
precede the optimization parameters in the OPAL input file.

@@

The initial number of individuals was fixed at 656, 
and the minimum number individuals in later generations was fixed at 328. 
These numbers \lsnote{were} in part based on the architecture of the KNL's. 
Since each simulation takes 8 cores, and there are 64 cores per KNL node, 
we wanted a large population size that would fit evenly on these resources.
